---
title: 'Introduction to Response Surface Analysis '
author: Johannes Karl
date: '2019-03-23'
slug: introduction-to-response-surface-analysis
categories:
  - R
  - Stats
tags:
  - R
  - RSA
  - plotting
image:
  caption: ''
  focal_point: ''
---

```{r loading data, include = F}
library(tidyverse)
setwd(here::here())
data_long <- read_csv("static/files/RSA.csv")
data_long$attr <- rowMeans(data_long[c("Attr_1", "Attr_2")]) 
data_long$val_cent <- data_long$Valence - 5
data_long$aro_cent <- data_long$Arousal - 5
data_long$val_cent2 <- data_long$val_cent * data_long$val_cent
data_long$aro_cent2 <- data_long$aro_cent * data_long$aro_cent
data_long$t_attr <- psych::rescale(data_long$attr,mean = 50, sd = 10)[[1]]
```

# The Theory

Response Suface Analysis (RSA) is a method from the wider field of Response Surface Methodologies (RSM). 

# Requirements

1. Commensurable Scales

     Commensurability explained as briefly as possible is that two scales are measured in the same
     units. In the world of Likert-scales this mostly means they are measured on the same number of
     points.
     
2. Centered on a Common Point

     There is a number of ways this can be done. The most straight forward is probably to subtract 
     scale midpoint from each participants score. This leaves participants with scores below the        theoretical midpoint with negative values and everyone else with positive values.
     
3. Participants can indicate high and low values on all scales
     
     This is more of a scale design than a analysis issue. Here it is important to consider if negatively worded items should need to be included in the scale or if the lowest level of a positively worded item is the lowest score a participant could indicate.
     
4. No Multicolliniearity
    
    In contrast to the previous requirement this one is actually testable. The easiest way of doing this would be to run a regression with all predictors and examine the Variance Inflation Factors (VIF). An example can be seen below where I predict attractiveness of a landscape from ratings of valence and arousal. valence and arousal are centered around the scale mid-point of 5. VIF values below **5** are generally considered acceptabily low to conclude that there is no mulitcolliniearity. 
    
    ```{r vif}
car::vif(lm(attr ~ val_cent + aro_cent, data = data_long))
```




5. High Power


# RSA How-To


```{r RSA}
library("RSA")
rsa_sqd <- RSA(t_attr ~ val_cent * aro_cent, data = data_long, model = "full")

summary(rsa_sqd)
# PLOT THE ESTIMATED MODEL
plot(rsa_sqd, legend=F, param=F,
     axes=c("LOC", "LOIC","PA1"), 
     project=c("LOC", "LOIC","PA1"), 
     xlab="Valence", ylab="Arousal", zlab="Attractiveness"
     )
```